Characteristic polynomials of p-adic matrices

### 2017-04-25 charpoly/charpoly.tex
Small fixes (typos, notations...)

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi(M)$ of 
an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
developed previously.  When $M$ is an integral matrix whose entries are 
all given at the same precision $O(p^N)$, we give a criterion (checkable 
within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
coefficient of $\chi(M)$ with more accuracy than $O(p^N)$.
In general, we provide two algorithms for determining the optimal 
precision of the coefficients of $\chi(M)$ and of their eigenvalues.
We provide evidence showing that classical algorithms do not reach
this optimal precision in general.
\end{abstract}

@@

%\begin{comment}

@@

\item \done{We agree that further experiments would be beneficial, but we are constrained by space (due to the 8 page limit) and time (due to the limited time to make revisions).  We note that we did perform the current experiment in higher dimension with similar results: dimension $8$ was chosen to be representative without the table taking up too much space in the paper.}

@@

\bigskip
%\end{comment}

@@

Let $M \in M_n(\Qp)$, with every entry given at precision $O(p^N)$.
The simplest approach for computing the characteristic polynomial of $M$
is to compute $\det(X-M)$ either using recursive row expansion or various division free
algorithms \citelist{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
with these methods.  First, they are slower than alternatives that allow division,
requiring $O((n+1)!)$, $\softO(n^5)$ and $\softO(n^{3 + \omega/2})$ operations in $K$.  Second,
while the lack of division implies that the result is accurate modulo $p^N$ as
long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.

@@

Instead, we separate the computation of the precision of $\chi_M$ from the computation
of an approximation to $\chi_M$.  Given some precision on $M$, we use \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}
to find the best possible precision for $\chi_M$.  The analysis of this precision is the subject
of much of this paper.  With this precision known, the actual calculation of $\chi_M$
may proceed by lifting $M$ to a temporarily higher precision and then using a sufficiently
stable algorithm (see Remark \ref{rem:lift_for_optimal}).

@@

This paper provides a theoretical and concrete study of the propagation 
of the precision during the computation of the characteristic polynomial 
of matrix defined over the $p$-adics. Thanks to the application of the 
general framework of differential precision developed in 
\cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a}, we know that the 
optimal precision of the characteristic polynomial $\chi_M$ of a matrix 
$M \in M_n(\mathbb{Q}_p)$ is controled by the adjugate $\adj(X{-}M).$
In this article, we provide:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\item Proposition \ref{prop:shortcom}: a short representation of
the matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
field (instead of $O(n^3)$ for a dense representation);
\item two algorithms to compute $\adj(X{-}M)$: the first one is 
division-free and performs $\softO(n^3)$ operations in $\Zp$ while 
the second one may perform divisions and has complexity $\tilde 
O(n^\omega)$;
\end{enumerate}
and deduce from this the following applications to the question
we are interested in:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\setcounter{enumi}{2}
\item Corollary \ref{cor:prec_gain}: a simple criterion to decide 
whether $\chi_M$ is defined at precision higher than the precision of 
$M$ (when $M$ lies in $M_n(\mathbb{Z}_p)$ and all its entries are given 
at the same precision);
\item two algorithms to compute the optimal precision on each 
coefficient of the characteristic polynomial of a matrix $M \in 
M_n(\Qp)$ (whose entries may be given at different precision): the first 
one is division-free and has complexity $\softO(n^3)$ while the second 
one may perform divisions but has complexity $\softO(n^\omega)$;
\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
algorithm to compute the optimal precision on each eigenvalue of $M$.
\end{enumerate}
Moreover, we provide evidence showing that classical algorithms for 
computing the characteristic polynomials do not reach the optimal 
precision in many contexts.

@@

One cannot represent
matrices $M \in M_n(K)$ exactly, but instead must specify an approximation
together with a measure of the approximation's accuracy; there are several ways to do so.
When using \emph{jagged} precision, on gives a precision on each entry,
representing the $(i,j)$ entry as $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
is called the \emph{absolute precision} and may depend on $(i,j)$.
When the $N_{i,j}$ are all equal, we say that the precision datum is 
\emph{flat}.

@@

When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may give
the precision of $\chi(M)$ precisely as
\begin{equation} \label{eq:mainlem}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
The equality in \eqref{eq:mainlem} justifies our use of the phrase \emph{optimal} precision
to denote the lattice precision $d\chi_M(H)$. To make the conditions on $H$ under
which \eqref{eq:mainlem} holds more precise, let
$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.

@@

For further discussion of using lattices to measure precision,
details on the computation of $d\chi_M$ in \eqref{eq:dchi}, and 
examples of additional precision in $\chi(M)$,
see \cite{caruso:17a}*{\S 2.4.3 and \S 3.2.2}

@@

For any $P \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
as the image of $d\chi$ at $PMP^{-1}$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the last column of
$\adj(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.

Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\adj(X-M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
where $\mu, d_\mu$ ranges over the eigenvalues and dimensions of the other Jordan blocks.
Since $\lambda$ occurs in two Jordan blocks,
$X - \lambda$ will divide every entry of $\adj(X-M)$ and $d\chi_M$ will not be surjective.
\end{proof}

@@

\begin{cor} \label{cor:prec_gain}
Suppose that $M \in \GL_n(\OK)$ is known with precision $O(\pi^m)$.
Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(\pi^m)$
if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
\end{cor}

@@

\noindent
{\bf Stability under multiplication by $X$.}
By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
However, when $M$ is given, $K_{< n}[X]$ is canonically isomorphic
to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
often be convenient to view $d \chi_M$ as an $K$-linear mapping
$M_n(K) \to K[X]/\chi_M(X)$.

@@

Proposition~\ref{prop:shortcom} shows that $\adj(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ entries to 
write down the matrices $P$ and $Q$. 
We shall see moreover in Section \ref{sec:diffFrob} that interesting
information can be read off of this short form $(\alpha, P, Q, 
\chi_M)$.

@@

It is not hard to prove that every matrix over a field admits
a Hessenberg form.
We prove here that we may compute an approximate Hessenberg form 
for any matrix known at finite (jagged) precision over $K$.
Moreover, we provide an exact change of basis matrix.

@@

\begin{prop} 
Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M$.
$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficient in $M.$
\end{prop}
\begin{proof}
Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
with the $\varepsilon$'s being units,
the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\Fq[X],$ $\Q[X]$ or adequate extension) of $\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}.$
Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
The rest is clear.
\end{proof}

@@

In this section, we prove that to compute the inverse of
a matrix over a CDVF $K$, the Smith normal form is precision-wise optimal in the flat-precision case.
We first recall the differential of matrix inversion.

@@

\begin{prop}
Let $\cond(M) = \val(\sigma_n(M))$.
If $dM$ is a flat precision of $O(\pi^m)$ on $M$ then $M^{-1}$
can be computed at precision $O(\pi^{m-2\cond(M)})$ by a \textbf{SNF} computation
and this lower-bound is optimal,
at least when $m$ is large.
\end{prop}
\begin{proof}
The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M).$
It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that $m-2\cond(M)$
can be obtained as the valuation of a coefficient of $du_M(dM)$
and the smallest that can be achieved this way for $dM$ in a flat precision lattice.
Hence the optimality of the bound given, for large 
$m$ \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.

Now, the computation of the Smith normal form was described in \cite{Vaccon-these}.
From $M$ known at flat precision $O(\pi^m),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
known at precision at least $O(\pi^{m-\cond(M)})$, with coefficients in $\OK$
and determinant in $\OK^\times$ realizing an Smith normal form of $M.$
There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
Since the smallest valuation occurring in $\Delta^{-1}$ is $-\cond(M),$
we see that $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ is known at precision at least $O(\pi^{m-2\cond(M)}),$
which concludes the proof.
\end{proof}

@@

In this section, we compute $\adj(X-H)$ for a Hessenberg matrix $H$
using the Smith normal form computation of the previous section.
The entries of $\adj(X-H)$ lie in $K[X]$, which is not a CDVF,
so we may not directly apply the methods of the previous section.
However, we may relate $\adj(X-H)$ to $\adj(1-XH)$, whose
entries lie in the CDVF $K\llp X\rrp$.  In this way, we compute $\adj(X-H)$
using an SNF method, with no division in $K$.

@@

The second ingredient we need is reciprocal polynomials.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $d \in \mathbb{N}$ and $f \in K[X]$ of degree at most $d.$ 
We define the reciprocal polynomial of order $d$ of $f$ as $f^{\rec,d}=X^d f \left( 1/X \right).$
Let $F \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
We denote by $F^{\rec,d}$ the matrix with $(F^{\rec,d})_{i,j} = (F_{i,j})^{\rec,d}$.
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
\adj(1-XM)^{\rec,n-1}=&\adj(X-M), \\
(\chi_M I_n)^{\rec,n}=&(1-XM) \adj(1-XM).\\
\end{eqnarray*}
\end{lem}
\begin{proof}
It all comes down to the following result:
let $F \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
then $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
Indeed, one can use multilinearity of the determinant on $X^d \det(F(1/X))$
to prove this result.
It directly implies the second part of the lemma; the first part follows
from the fact that the entries of $\adj(X-M)$ and of $\adj(1-XM)$
are determinants of size $n-1$.
\end{proof}

@@

\begin{theo}
Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix with
coefficient-wise precision $O(\pi^M).$
Then, using Algorithm 2, one can compute $\adj (X -H)$
and $\chi_H$ in 
$\softO (n^3)$ operations in $\OK$ at the precision given by $H.$
\end{theo}
\begin{proof}
First, the operations of the lines 2 and 3 use $\softO (n^3)$ operations
in $\OK$ at the precision given by $H.$
Indeed, since $H$ is an approximate Heisenberg matrix, when we use $U_{i,i}$ as pivot
the only other nonzero coefficient in its column is $U_{i+1,i}$.
As a consequence, when performing this column-pivoting, only two rows ($i$ and
$i+1$) lead to operations in $\OK \llb X \rrb$ other than checking precision.
Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
Following line 1, the computation of $Q$ is done by operations on rows, starting from the identity matrix.
The order in which the entries of $U$ are cleared implies that $Q$ is just filled in as an upper triangular matrix:
no additional operations in $\OK\llb X\rrb$ are required. Thus the total cost
for lines 2 and 3 is indeed $\softO (n^3)$ operations.

For lines 4 and 5, there are only $n-1$ eliminations, resulting in a $\softO (n^2)$ cost
for the computation of $U.$ Rather than actually construct $P$, we just track the eliminations
performed in order to do the corresponding row operations on $Q$, since we only need the product $P \times Q$.

Line 6 is in $\softO (n^2)$ and 7 in $\softO (n^3).$

Thanks to the fact that the $P$ only corresponds to the product of $n-1$ 
shear matrices, the
product on line 8 is in $\softO (n^3).$
We emphasize that no division has been done.
Line 9 is costless, and the result is proven.
\end{proof}
\begin{rem}
If $M \in M_n(K)$ does not have coefficients in $\OK,$
we may apply Algorithms 1 and 2 to $p^v M \in M_n(\OK)$
in $\softO (n^3)$ operations in $\OK$, and then divide
the coefficient of $X^k$ in the resulting polynomial by $p^{kv}$.
\end{rem}

@@

In this section, we combine Proposition \ref{prop:shortcom}
with Algorithm 2 to compute the adjugate of $X-M$
when $\chi_M$ is squarefree.   Note that, under
the assumption that $d\chi_M$ is surjective,
$\chi_M$ is squarefree if and only if $M$ is diagonalizable.  The result is
the following $\softO(n^3)$ algorithm, where the only
divisions are for gcd and modular inverse computations.

@@

The idea is to work modulo $X-\lambda$
for $\lambda$ a root of $\chi (M)$ (in an algebraic closure)
and then apply Chinese Remainder Theorem.
The goal of the Step $2$ is to ensure the first row of $B$ contains an
invertible entry modulo $\chi_M$.
Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
affine hyperplane so that $\row(B,1) \mod (X-\lambda)$ is a non-zero vector.
Hence we need only avoid a finite union of affine hyperplanes
in order for $\row(B,1) \mod \chi (M)$ to contain an invertible coefficient.

@@

The algorithm designed in the previous section computes the differential 
$d \chi_M$ of $\chi$ at a given matrix $M \in M_n(K)$ for a cost of 
$\softO (n^3)$ operations in $K$. This seems to be optimal given that the 
(naive) size of the $d \chi_M$ is $n^3$: it is a matrix of size $n
\times n^2$. It turns out however that improvements are still possible!
Indeed, thanks to Proposition~\ref{prop:shortcom}, the matrix of 
$d \chi_M$ admits a compact form which can be encoded using only $O(n^2)$ 
coefficients. The aim of this short section is to design a fast 
algorithm (with complexity $\softO(n^\omega)$) for computing this short 
form. The price to pay is that divisions in $K$ appear, which can be an 
issue regarding to precision in particular cases.
In this section, we only estimate the number of operations in $K$
and not their behavior on precision.

@@

\begin{theo}
\label{thm:compute_shortcom}
Given $M \in M_n(K)$ with $d \chi_M$ surjective,
one can compute $\alpha, P, Q, \chi_M) \in K[X]$ with
$\softO(n^\omega)$ operations in $K$ such that
$\adj(X{-}N) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$.
\end{theo}

@@

In the previous Sections, \ref{sec:diffHess} and \ref{sec:diffFrob},
we have proposed algorithms to obtain the adjugate of 
$X-M.$ Our motivation for these computations is to then
be able to understand what is the optimal precision on $\chi_M.$
In this section, we provide some answers to this question,
along with numerical evidence.
We also show that it is then possible to derive 
optimal precision of eigenvalues of $M.$

@@

\begin{enumerate}[$\bullet$]
\renewcommand{\itemsep}{0pt}
\item the optimal precision on the $k$-th coefficient of the 
characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}.
Note that the method for computing $\adj(X-M)$ is irrelevant
since we only record the precision loss, not the runtime.
\item in the capped relative model\footnote{Each 
coefficient carries its own precision which is updated after each 
elementary arithmetic operation.},
the precision on the $k$-th coefficient of the 
characteristic polynomial of $M$ is computed \emph{via} the call:

>>>>

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi_M$ of 
an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
developed previously.  When $M$ is an integral matrix whose entries are 
all given at the same precision $O(p^N)$, we give a criterion (checkable 
within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
coefficient of $\chi_M$ with more accuracy than $O(p^N)$.
In general, we provide two algorithms for determining the optimal 
precision of the coefficients of $\chi_M$ and of their eigenvalues.
We provide evidence showing that classical algorithms do not reach
this optimal precision in general.
\end{abstract}

@@

\begin{comment}

@@

\item \done{We agree that further experiments would be beneficial, but we are constrained by space (due to the 8 page limit) and time (due to the limited time to make revisions).  We note that we did perform the current experiment in higher dimension with similar results: dimension $9$ was chosen to be representative without the table taking up too much space in the paper.}

@@

\bigskip
\end{comment}

@@

Let $M \in M_n(\Qp)$, with every entry given at precision $O(p^N)$.
The simplest approach for computing the characteristic polynomial of $M$
is to compute $\det(X{-}M)$ either using recursive row expansion or various division free
algorithms \citelist{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
with these methods.  First, they are slower than alternatives that allow division,
requiring $O((n+1)!)$, $\softO(n^5)$ and $\softO(n^{3 + \omega/2})$ operations in $K$.  Second,
while the lack of division implies that the result is accurate modulo $p^N$ as
long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.

@@

Instead, we separate the computation of the precision of $\chi_M$ from the computation
of an approximation to $\chi_M$.  Given some precision on $M$, we use the theory developed in \cite{caruso-roe-vaccon:14a}
to find the best possible precision for $\chi_M$.  The analysis of this precision is the subject
of much of this paper.  With this precision known, the actual calculation of $\chi_M$
may proceed by lifting $M$ to a temporarily higher precision and then using a sufficiently
stable algorithm (see Remark \ref{rem:lift_for_optimal}).

@@

This paper provides a theoretical and concrete study of the propagation 
of the precision during the computation of the characteristic polynomial 
of matrix defined over the $p$-adics. Thanks to the application of the 
general framework of differential precision developed in 
\cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a}, we know that the 
optimal precision of the characteristic polynomial $\chi_M$ of a matrix 
$M \in M_n(\mathbb{Q}_p)$ is controled by the adjugate $\adj(X{-}M).$
In this article, we provide:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\item Proposition \ref{prop:shortcom}: a short representation of
the matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
field (instead of $O(n^3)$ for a dense representation);
\item two algorithms to compute $\adj(X{-}M)$: the first one is 
division-free and performs $\softO(n^3)$ operations in $\Zp$ while 
the second one may perform divisions and has complexity $\softO(n^\omega)$;
\end{enumerate}
and deduce from this the following applications to the question
we are interested in:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\setcounter{enumi}{2}
\item Corollary \ref{cor:prec_gain}: a simple criterion to decide 
whether $\chi_M$ is defined at precision higher than the precision of 
$M$ (when $M$ lies in $M_n(\mathbb{Z}_p)$ and all its entries are given 
at the same precision);
\item two algorithms to compute the optimal precision on each 
coefficient of the characteristic polynomial of a matrix $M \in 
M_n(\Qp)$ (whose entries may be given at different precision): the first 
one is division-free and has complexity $\softO(n^3)$ (operations in
$\Zp$) while the second 
one may perform divisions but has complexity $\softO(n^\omega)$;
\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
algorithm to compute the optimal precision on each eigenvalue of $M$.
\end{enumerate}
Moreover, we provide evidence showing that classical algorithms for 
computing the characteristic polynomials do not reach the optimal 
precision in many contexts.

@@

One cannot represent
matrices $M \in M_n(K)$ exactly, but instead must specify an approximation
together with a measure of the approximation's accuracy; there are several ways to do so.
When using \emph{jagged} precision, one gives a precision on each entry,
representing the $(i,j)$ entry as $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
is called the \emph{absolute precision} and may depend on $(i,j)$.
When the $N_{i,j}$ are all equal, we say that the precision datum is 
\emph{flat}.

@@

When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may give
the precision of $\chi_M$ precisely as
\begin{equation} \label{eq:mainlem}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
The equality in \eqref{eq:mainlem} justifies our use of the phrase \emph{optimal} precision
to denote the lattice precision $d\chi_M(H)$. To make the conditions on $H$ under
which \eqref{eq:mainlem} holds more precise, let
$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.

@@

For further discussion of using lattices to measure precision,
details on the computation of $d\chi_M$ in \eqref{eq:dchi}, and 
examples of additional precision in $\chi_M$,
see \cite{caruso:17a}*{\S 2.4.3 and \S 3.2.2}.

@@

For any $P \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
as the image of $d\chi$ at $PMP^{-1}$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the last column of
$\adj(X{-}M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.

Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\adj(X{-}M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
where $\mu, d_\mu$ ranges over the eigenvalues and dimensions of the other Jordan blocks.
Since $\lambda$ occurs in two Jordan blocks,
$X - \lambda$ will divide every entry of $\adj(X{-}M)$ and $d\chi_M$ will not be surjective.
\end{proof}

@@

\begin{cor} \label{cor:prec_gain}
Suppose that $M \in \GL_n(\OK)$ is known with flat precision $O(\pi^N)$.
Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(\pi^N)$
if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
\end{cor}

@@

\noindent
{\bf Stability under multiplication by $X$.}
By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
However, when $M$ is given, $K_{< n}[X]$ is canonically isomorphic
to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
often be convenient to view $d \chi_M$ as a $K$-linear mapping
$M_n(K) \to K[X]/\chi_M(X)$.

@@

Proposition~\ref{prop:shortcom} shows that $\adj(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$: the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ entries to 
write down the matrices $P$ and $Q$. 
We shall see moreover in Section \ref{sec:diffFrob} that interesting
information can be read off of this short form $(\alpha, P, Q, 
\chi_M)$.

@@

It is not hard to prove that every matrix over a field admits
a Hessenberg form.
We prove here that we may compute an approximate Hessenberg form 
for any matrix known at finite jagged precision over $K$.
Moreover, we provide an exact change of basis matrix.

@@

\begin{prop} 
Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M$.
$P$ is exact 
%over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, 
and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficient in $M.$
\end{prop}
\begin{proof}
%Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
with the $\varepsilon$'s being units,
the corresponding coefficient of the corresponding shear matrix is 
%the lift(in $\mathbb{Z}, $  $\Fq[X],$ $\Q[X]$ or adequate extension) of 
any exact element of $K$ congruent to 
$\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}$.
Exactness follows directly. 
%Over other fields, we can not lift, but the computations are still valid.
The rest is clear.
\end{proof}

@@

In this section, we prove that to compute the inverse of
a matrix over $K$, the Smith normal form is precision-wise optimal in the flat-precision case.
We first recall the differential of matrix inversion.

@@

\begin{prop}
Let $\cond(M) = \val(\sigma_n(M))$.
If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
can be computed at precision $O(\pi^{N-2\cond(M)})$ by a \textbf{SNF} computation
and this lower-bound is optimal,
at least when $N$ is large.
\end{prop}

\begin{proof}
The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M)$.
%It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that 
Consequently $N - 2\cond(M)$
can be obtained as the valuation of a coefficient of $du_M(dM)$
and it is the smallest that can be achieved this way for $dM$ in a flat precision lattice.
Hence the optimality of the bound given, for large 
$N$ \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.

Now, the computation of the Smith normal form was described in \cite{Vaccon-these}.
From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
known at precision at least $O(\pi^{N-\cond(M)})$, with coefficients in $\OK$
and determinant in $\OK^\times$ realizing an Smith normal form of $M.$
There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
Since the smallest valuation occurring in $\Delta^{-1}$ is $-\cond(M),$
we see that $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ is known at precision at least $O(\pi^{N-2\cond(M)}),$
which concludes the proof.
\end{proof}

@@

In this section, we compute $\adj(X-H)$ for a Hessenberg matrix $H$
using the Smith normal form computation of the previous section.
The entries of $\adj(X-H)$ lie in $K[X]$, which is not a complete
discrete valuation field,
so we may not directly apply the methods of the previous section.
However, we may relate $\adj(X-H)$ to $\adj(1-XH)$, whose
entries lie in $K\llp X\rrp$.  In this way, we compute $\adj(X-H)$
using an SNF method, with no division in $K$.

@@

The second ingredient we need is reciprocal polynomials.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $d \in \mathbb{N}$ and $f \in K[X]$ of degree at most $d.$ 
We define the reciprocal polynomial of order $d$ of $f$ as $f^{\rec,d}=X^d f \left( 1/X \right).$
Let $F \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
We denote by $F^{\rec,d}$ the matrix with $(F^{\rec,d})_{i,j} = (F_{i,j})^{\rec,d}$.
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{align*}
\adj(1-XM)^{\rec,n-1}&=\adj(X-M), \\
(\chi_M I_n)^{\rec,n}&=(1-XM) \adj(1-XM).\\
\end{align*}
\end{lem}
\begin{proof}
It all comes down to the following result:
let $F \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
then $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
%Indeed, one can use multilinearity of the determinant on $X^d \det(F(1/X))$ to prove this result.
It directly implies the second part of the lemma; the first part follows
from the fact that the entries of $\adj(X-M)$ and of $\adj(1-XM)$
are determinants of size $n-1$.
\end{proof}

@@

\begin{theo}
Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix with
flat precision $O(\pi^N).$
Then, using Algorithm 2, one can compute $\adj (X{-}H)$
and $\chi_H$ in 
$\softO (n^3)$ operations in $\OK$ at precision $O(\pi^N).$
\end{theo}
\begin{proof}
First, the operations of the lines 2 and 3 use $\softO (n^3)$ operations
in $\OK$ at precision $O(\pi^N).$
Indeed, since $H$ is an approximate Heisenberg matrix, when we use $U_{i,i}$ as pivot
the only other nonzero coefficient in its column is $U_{i+1,i}$.
As a consequence, when performing this column-pivoting, only two rows ($i$ and
$i+1$) lead to operations in $\OK \llb X \rrb$ other than checking precision.
Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
Following line 1, the computation of $Q$ is done by operations on rows, starting from the identity matrix.
The order in which the entries of $U$ are cleared implies that $Q$ is just filled in as an upper triangular matrix:
no additional operations in $\OK\llb X\rrb$ are required. Thus the total cost
for lines 2 and 3 is indeed $\softO (n^3)$ operations.

For lines 4 and 5, there are only $n{-}1$ eliminations, resulting in a $\softO (n^2)$ cost
for the computation of $U.$ Rather than actually constructing $P$, we just track the eliminations
performed in order to do the corresponding row operations on $Q$, since we only need the product $P \times Q$.

Line 6 is in $\softO (n^2)$ and line 7 in $\softO (n^3).$

Thanks to the fact that the $P$ only corresponds to the product of $n{-}1$ 
shear matrices, the
product on line 8 is in $\softO (n^3).$
We emphasize that no division has been done.
Line 9 is costless, and the result is proven.
\end{proof}
\begin{rem}
If $M \in M_n(K)$ does not have coefficients in $\OK,$
we may apply Algorithms 1 and 2 to $p^v M \in M_n(\OK)$
in $\softO (n^3)$ operations in $\OK$, and then divide
the coefficient of $X^k$ in the resulting polynomial by $p^{kv}$.
\end{rem}

@@

In this section, we combine Proposition \ref{prop:shortcom}
with Algorithm 2 to compute the adjugate of $X{-}M$
when $\chi_M$ is squarefree.   Note that, under
the assumption that $d\chi_M$ is surjective,
$\chi_M$ is squarefree if and only if $M$ is diagonalizable.  The result is
the following $\softO(n^3)$ algorithm, where the only
divisions are for gcd and modular inverse computations.

@@

The idea is to work modulo $X-\lambda$
for $\lambda$ a root of $\chi_M$ (in an algebraic closure)
and then apply Chinese Remainder Theorem.
The goal of the Step $2$ is to ensure the first row of $B$ contains an
invertible entry modulo $\chi_M$.
Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
affine hyperplane so that $\row(B,1) \mod (X-\lambda)$ is a non-zero vector.
Hence we need only avoid a finite union of affine hyperplanes
in order for $\row(B,1) \mod \chi (M)$ to contain an invertible coefficient.

@@

The algorithm designed in the previous section computes the differential 
$d \chi_M$ of $\chi$ at a given matrix $M \in M_n(K)$ for a cost of 
$\softO (n^3)$ operations in $K$. This seems to be optimal given that 
the (naive) size of the $d \chi_M$ is $n^3$: it is a matrix of size $n 
\times n^2$. It turns out however that improvements are still possible! 
Indeed, thanks to Proposition~\ref{prop:shortcom}, the matrix of $d 
\chi_M$ admits a compact form which can be encoded using only $O(n^2)$ 
coefficients. The aim of this short section is to design a fast 
algorithm (with complexity $\softO(n^\omega)$ operations in $K$) for 
computing this short form. The price to pay is that divisions in $K$ 
appear, which can be an issue regarding to precision in particular 
cases. In this section, we only estimate the number of operations in $K$ 
and not their behavior on precision.

@@

\begin{theo}
\label{thm:compute_shortcom}
Given $M \in M_n(K)$ with $d \chi_M$ surjective,
one can compute $(\alpha, P, Q, \chi_M) \in K[X]$ with
$\softO(n^\omega)$ operations in $K$ such that
$\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$.
\end{theo}

@@

In the previous Sections, \ref{sec:diffHess} and \ref{sec:diffFrob},
we have proposed algorithms to obtain the adjugate of 
$X{-}M.$ Our motivation for these computations is to then
be able to understand what is the optimal precision on $\chi_M.$
In this section, we provide some answers to this question,
along with numerical evidence.
We also show that it is then possible to derive 
optimal precision of eigenvalues of $M.$

@@

\begin{enumerate}[$\bullet$]
\renewcommand{\itemsep}{0pt}
\item the optimal precision on the $k$-th coefficient of the 
characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}.
(Note that the method for computing $\adj(X{-}M)$ is irrelevant
since we only record the precision loss, not the runtime.)
\item in the capped relative mode (each 
coefficient carries its own precision which is updated after each 
elementary arithmetic operation),
the precision on the $k$-th coefficient of the 
characteristic polynomial of $M$ is computed \emph{via} the call:
